Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-20 aistats.tex
Prove gaussian logpart is not self-concordant

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177]{nesterov2003introductory} % TODO example 4.1.1.4
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $\mu_0, n_0$ spanning $[10^{-2}, 10^2]$. We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
	}
	\label{fig:optimal_n0}
\end{figure}

@@

\rlp{prove that $A$ is not self -concordant, while, $A^*$ is thanks to Nesterov.}
Maybe plot both of them.
Also plot the mirror maps.
And plot gaussian trajectories.

>>>>

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\  % TODO remove factor 2
	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177]{nesterov2003introductory} % TODO add citation to example 4.1.1.4
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

In this section, we prove all the results mentioned in \S\ref{ssec:gaussian-variance}, and add some context and experimental observations.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$. The expectation was estimated with $10^4$ draws for each value of $n_0, \mu_0$ We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
	}
	\label{fig:optimal_n0}
\end{figure}

@@

In this section, we prove that for a Gaussian, the entropy is self-concordant while the log-partition function is not.
We also give complementary illustrations on the Gaussian mirror map and MLE.

The Gaussian log-partition function and entropy are up to constants
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
These functions and their domains they are represented in Figure \TODO{}.

\begin{definition}[Self-concordance]
A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}$
\end{definition}

\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(x)} \leq C \norm{v}_2 g''(x)$ for some $C\geq 0$
\end{definition}

\paragraph{Entropy is Self-Concordant,}
\citet[p.177]{nesterov2003introductory} proves that logarithmic barriers for a second order region are self-concordant. Adopting their notation, $\conj$ fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$.

\paragraph{Log-partition is not Self-Concordant}
Let us fix a straight line with $\nat_1 \neq 0$, e.g. $g(\nat_2) := A(\nat_1,\nat_2)$. Then
\alignn{
g'(\nat_2) = \frac{\partial A}{\partial\nat_2}(\nat) 
&= +\frac{\nat_1^2}{4 \nat_2^2} + \frac{1}{2\nat_2} \\
g''(\nat_2) = \frac{\partial^2 A}{(\partial\nat_2)^2}(\nat) 
&= -\frac{\nat_1^2}{2 \nat_2^3} - \frac{1}{2\nat_2^2} 
=  \frac{1}{2\nat_2^2}\paren{\frac{\nat_1^2}{-\nat_2^2} - 1} \\ 
g'''(\nat_2) = \frac{\partial^3 A}{(\partial\nat_2)^3}(\nat) 
&= +\frac{3\nat_1^2}{2 \nat_2^4} + \frac{1}{\nat_2^3}
= \frac{1}{2\nat_2^3}\paren{2 - 3 \frac{\nat_1^2}{- \nat_2^2}}
}
Setting $\nat_2 = - \nat_1^2$ yields $g''(\nat_2) = 0$ and $g'''(\nat_2) \neq 0$, meaning the second derivative cannot dominate the third derivative for any choice of constant. The log-partition is neither self concordant nor pseudo-self-concordant.
